Analysis-ready VCF at Biobank scale using Zarr

### 2025-01-28 paper.tex

%%% Use the \authfn to add symbols for additional footnotes, if any.
% 1 is reserved for correspondence emails; then continuing with 2 etc for contributions.
% First author
\author[1,\authfn{1},\authfn{3}]{Eric Czech} % https://orcid.org/0000-0002-4254-4255
\author[2,3\authfn{1}]{Timothy R. Millar} % https://orcid.org/0000-0002-5142-8811
\author[4,\authfn{1}]{Will Tyler}
\author[5,\authfn{1}]{Tom White}

@@

% Senior
\author[1,\authfn{2},\authfn{3}]{Jeff Hammerbacher} % https://orcid.org/0000-0001-6596-8563
\author[10,\authfn{2},\authfn{4}]{Jerome Kelleher} % https://orcid.org/0000-0002-7894-5253

\affil[1]{Open Athena AI Foundation}
\affil[2]{The New Zealand Institute for Plant \& Food Research Ltd, Lincoln,
New Zealand}
\affil[3]{Department of Biochemistry, School of Biomedical Sciences, University of Otago, Dunedin, New Zealand}
\affil[4]{Independent researcher}
\affil[5]{Tom White Consulting Ltd.}
\affil[6]{Our Future Health, Manchester, UK.}
\affil[7]{Program in Medical and Population Genetics, Broad Institute of MIT and Harvard, Cambridge, Massachusetts 02142, USA}
\affil[8]{Analytic and Translational Genetics Unit, Massachusetts General
Hospital, Boston, Massachusetts 02114, USA}
\affil[9]{NVIDIA Ltd, Reading, UK}
\affil[10]{Big Data Institute, Li Ka Shing Centre for Health Information and Discovery,
University of Oxford, UK}
\affil[11]{Novo Nordisk Foundation Center for Genomic Mechanisms of Disease, Broad Institute of MIT and Harvard, Cambridge, Massachusetts 02142, USA}
\affil[12]{Wellcome Sanger Institute}
\affil[13]{Genomics England}
\affil[14]{Department of Cell and Molecular Biology, National
  Bioinformatics Infrastructure Sweden, Science for Life Laboratory,
  Uppsala University, Uppsala, Sweden}
\affil[15]{School of Computer Science, McGill University, Montreal, QC, Canada}

%%% Author Notes
\authnote{\authfn{1}Joint first author.}
\authnote{\authfn{2}Joint senior author.}
\authnote{\authfn{3}Work partially performed at Related Sciences.}
\authnote{\authfn{4}jerome.kelleher@bdi.ox.ac.uk}

@@

% We're not pretending is a solution to all problems and
% up front about weaknesses of the model.
Zarr provides pragmatic solutions to some of the more pressing
problems facing the analysis of large-scale genetic variation
data, but it is not a panacea. Firstly,
any dataset containing a variant with a large number of alleles
(perhaps due to indels) will cause problems because the
dimensions of fields are determined by their \emph{maximum}
dimension among all variants. In particular this is problematic
for fields like PL in which the dimension depends
quadratically on the number of alleles.
While the local alleles representation~\cite{poterba2025scalable}
is a substantial improvement (see All of Us and Norway Spruce
case studies) it may be lossy (see Methods). 
Secondly, the design of
VCF Zarr emphasises efficiency of analysis for a fixed
dataset, and does not consider how samples (and the
corresponding novel variants) should be added.
Thirdly, Zarr works best for numerical data of a fixed
dimension, and therefore may not suitable for representing
the unstructured data often included in VCF INFO fields.

@@

An important aspect of VCF Zarr is that field dimensions are
global and fixed, and defined as the maximum across all rows.
Continuing the example above, the third dimension of the array
is the maximum number of alternate alleles across \emph{all}
variants. For variants at which there are less than the maximum
number of alternative alleles, the third dimension of the
\texttt{call\_AD} array
is padded with a sentinel value (-2 for integers and a specific
non-signalling NaN for floats). This is problematic in situations
where some sites have a large number of alleles (e.g., 95 in the All
of Us case study above), and 
in particular for fields that have a quadratic dependency on the number of alleles
(Number=G) such as PL. We use the local alleles
encoding~\cite{poterba2025scalable} to reduce the storage 
requirements for such fields (see the next section for 
implementation details).

@@

The implementation of the local alleles data reduction technique 
in \texttt{vcf2zarr} follows the approach of Poterba 
et al~\cite{poterba2025scalable}.
Specifically, we identify the alleles in use for each call by
examining the genotypes and then store the LA field,
listing the (maximum of ploidy) unique alleles observed as indexes into the 
global list for that variant.
Using the LA field we then compute derived fields such as 
AD and PL in terms of the local alleles during Zarr encoding
(see e.g.\ the \texttt{call\_LAD} field in Table~\ref{tab-aou-data}
and
\texttt{call\_LPL} field in Table~\ref{tab-spruce-data}).
Note that our approach (along with Poterba et al)
differs from the details in the VCF 4.5 specification~\citep{vcf4_5},
which defines the LAA field containing indexes of ALT 
alleles only. Explicitly listing the alleles observed leads
to simpler code and means we do not need to refer back to the 
genotypes to distinguish missing data from homozygous reference 
calls. Note also that the local alleles encoding may be lossy,
as only information relevant to the chosen alleles for 
each call is retained.

\subsection{Choosing default compressor settings}
To inform the choice of compression settings across different fields
in VCF data, we analysed their effect on compression ratio on
recent high-coverage WGS data from the 1000 Genomes
project~\citep{byrska2022high}. We began by downloading
the first 100,000 lines of the VCF
for chromosome 22 (giving a 1.1GiB compressed VCF)
and converted to Zarr using
\texttt{vcf2zarr} with default settings. We then systematically
examined the effects of varying chunk sizes and compressor settings
on the compression ratio for call-level fields. (
We excluded \texttt{call\_PL} from this analysis as this analysis predated 
our implementation of local alleles.)

@@

\subsection{AWS parallel Zarr prototype}
This section describes the methodology used for the AWS parallel Zarr
throughput measurements in Fig~\ref{fig-s3-throughput}
and Fig~\ref{fig-s3-network-throughput}, and timing measurements
detailed in the cloud computing results.
These measurements were run on EC2 instances using jupyter notebooks
hosted on Genomics England's Lifebit CloudOS platform. VCF Zarr chunks
were hosted in a standard profile S3 bucket. Both instance and bucket were
in the same availability zone in the eu-west-2 region. Each
of the parallel Python processes used the aioboto3 library (version 13.2.0) to
fetch chunks concurrently from S3. The underlying aiohttp library
limits the concurrent fetching to 100 requests per process. For decompressing
and computing af-dist the same architecture was used but as fetched chunks
became available in memory in each process, they were decompressed and
computed sequentially. Decompression was performed by the blosc library
without the use of zarr.

>>>>

%%% Use the \authfn to add symbols for additional footnotes, if any.
% 1 is reserved for correspondence emails; then continuing with 2 etc for contributions.
% First author
\author[1,\authfn{1}]{Eric Czech} % https://orcid.org/0000-0002-4254-4255
\author[2,3\authfn{1}]{Timothy R. Millar} % https://orcid.org/0000-0002-5142-8811
\author[4,\authfn{1}]{Will Tyler}
\author[5,\authfn{1}]{Tom White}

@@

% Senior
\author[1,\authfn{2}]{Jeff Hammerbacher} % https://orcid.org/0000-0001-6596-8563
\author[10,\authfn{2},\authfn{3}]{Jerome Kelleher} % https://orcid.org/0000-0002-7894-5253

\affil[1]{Related Sciences}
\affil[2]{The New Zealand Institute for Plant \& Food Research Ltd, Lincoln,
New Zealand}
\affil[3]{Department of Biochemistry, School of Biomedical Sciences, University of Otago, Dunedin, New Zealand}
\affil[4]{Independent researcher}
\affil[5]{Tom White Consulting Ltd.}
\affil[6]{Our Future Health, Manchester, UK.}
\affil[7]{Program in Medical and Population Genetics, Broad Institute of MIT and Harvard, Cambridge, Massachusetts 02142, USA}
\affil[8]{Analytic and Translational Genetics Unit, Massachusetts General
Hospital, Boston, Massachusetts 02114, USA}
\affil[9]{NVIDIA Ltd, Reading, UK}
\affil[10]{Big Data Institute, Li Ka Shing Centre for Health Information and Discovery,
University of Oxford, UK}
\affil[11]{Novo Nordisk Foundation Center for Genomic Mechanisms of Disease, Broad Institute of MIT and Harvard, Cambridge, Massachusetts 02142, USA}
\affil[12]{Wellcome Sanger Institute}
\affil[13]{Genomics England}
\affil[14]{Department of Cell and Molecular Biology, National
  Bioinformatics Infrastructure Sweden, Science for Life Laboratory,
  Uppsala University, Uppsala, Sweden}
\affil[15]{School of Computer Science, McGill University, Montreal, QC, Canada}

%%% Author Notes
\authnote{\authfn{1}Joint first author.}
\authnote{\authfn{2}Joint senior author.}
\authnote{\authfn{3}jerome.kelleher@bdi.ox.ac.uk}

@@

% We're not pretending is a solution to all problems and
% up front about weaknesses of the model.
Zarr provides pragmatic solutions to some of the more pressing
problems facing the analysis of large-scale genetic variation
data, but it is not a panacea. Firstly,
any dataset containing a variant with a large number of alleles
(perhaps due to indels) will cause problems because the
dimensions of fields are determined by their \emph{maximum}
dimension among all variants. In particular this is problematic
for fields like PL in which the dimension depends
quadratically on the number of alleles.
While the local alleles representation~\cite{poterba2024scalable}
is a substantial improvement (see All of Us and Norway Spruce
case studies) it may be lossy (see Methods). 
Secondly, the design of
VCF Zarr emphasises efficiency of analysis for a fixed
dataset, and does not consider how samples (and the
corresponding novel variants) should be added.
Thirdly, Zarr works best for numerical data of a fixed
dimension, and therefore may not suitable for representing
the unstructured data often included in VCF INFO fields.

@@

An important aspect of VCF Zarr is that field dimensions are
global and fixed, and defined as the maximum across all rows.
Continuing the example above, the third dimension of the array
is the maximum number of alternate alleles across \emph{all}
variants. For variants at which there are less than the maximum
number of alternative alleles, the third dimension of the
\texttt{call\_AD} array
is padded with a sentinel value (-2 for integers and a specific
non-signalling NaN for floats). While this is not a problem
in practice for datasets in which all four bases are observed,
it is a substantial issue for
fields that have a quadratic dependency on the number of alleles
(Number=G) such as PL. Such fields are already known to cause
significant problems, and the ``local alleles'' proposal
provides an elegant solution~\cite{poterba2024scalable}.
As this approach is on a likely path to
standardisation~\cite{danecek2021twelve}, we plan to
include support in later versions of VCF Zarr.

@@

\subsection{Local alleles}
[TODO Implementation of local alleles in vcf2zarr]

\subsection{Choosing default compressor settings}
To inform the choice of compression settings across different fields
in VCF data, we analysed their effect on compression ratio on
recent high-coverage WGS data from the 1000 Genomes
project~\citep{byrska2022high}. We began by downloading
the first 100,000 lines of the VCF
for chromosome 22 (giving a 1.1GiB compressed VCF)
and converted to Zarr using
\texttt{vcf2zarr} with default settings. We then systematically
examined the effects of varying chunk sizes and compressor settings
on the compression ratio for call-level fields. We excluded call\_PL
from this analysis as it requires conversion to a ``local alleles''
encoding~\citep{poterba2024scalable} to be efficient,
which is planned for implementation in a future
version of \texttt{vcf2zarr}.

@@

\subsection{AWS parallel Zarr prototype}
[TODO Implementation of prototype Zarr implementaion]
